---
title: "CITS4009 Project - Exploratory Data Analysis and Predictive Modeling"
author: "Binghan CHEN(23965953)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

The objective of this project is to perform exploratory data analysis (EDA) and predictive modeling using the "Countries and Death Causes" dataset provided by the World Health Organisation (WHO). This dataset includes various health-related metrics for different countries over several years. The goal is to uncover insights from the data and build predictive models to understand the factors influencing death rates.

# Section 1: Exploratory Study of the Data

## Data Overview

### Loading the Dataset

```{r load-data}

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)

# Read the dataset (replace 'path_to_your_file.csv' with your actual file path)
data <- read.csv("Countries_and_Death_Causes.csv", stringsAsFactors = FALSE)

# Display the first few rows of the dataset
head(data)

```

### Understanding the Structure of the Data

```{r}
str(data)
```

### Summary Statistics

```{r}
summary(data)
```

## Data Cleaning

### Checking for Missing Values

```{r}
# Function to count missing values in each column
count_missing <- function(df) {
  sapply(df, function(col) sum(is.na(col)))
}

nacounts <- count_missing(data) 
hasNA <- which(nacounts > 0) 
nacounts[hasNA]
```

### Handling Missing Values
Based on the missing values identified, we will decide how to handle them—either by removing rows with missing values or by imputing them.

```{r}
# Remove rows with any missing values
data_clean <- na.omit(data)

# Verify that missing values are handled
count_missing(data_clean)
```

### Handling Data Anomalies
Identify and handle any outliers or anomalies in the data that may affect the analysis.

```{r}
# Example: Detecting outliers in 'Outdoor air pollution' using boxplot
ggplot(data_clean, aes(y = Outdoor.air.pollution)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Boxplot of Outdoor Air Pollution", y = "Outdoor Air Pollution") +
  theme_minimal()
```
Describe any actions taken to handle anomalies, such as removing outliers or transforming data.
Actions Based on Outliers:

Remove Outliers: If outliers are errors or anomalies not relevant to the analysis.
Transform Data: Apply transformations like log or square root to reduce the impact of outliers.
{r
Copy code
# Example: Removing outliers in 'Outdoor air pollution' beyond 1.5*IQR
Q1 <- quantile(data_clean$Outdoor.air.pollution, 0.25)
Q3 <- quantile(data_clean$Outdoor.air.pollution, 0.75)
IQR <- Q3 - Q1

# Define lower and upper bounds
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Filter out the outliers
data_clean <- data_clean %>%
  filter(Outdoor.air.pollution >= lower_bound & Outdoor.air.pollution <= upper_bound)

# Verify the removal
ggplot(data_clean, aes(y = Outdoor.air.pollution)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Boxplot of Outdoor Air Pollution (After Outlier Removal)", y = "Outdoor Air Pollution") +
  theme_minimal()

### Final Cleaned Dataset
After handling missing values and anomalies, we have a clean dataset ready for further analysis.

```{r
Copy code
# Display the first few rows of the cleaned dataset
head(data_clean)
```

Provide a brief summary of the cleaning steps and their impact on the dataset.

## Visualizations Transformations 
This section will include various visualizations and analyses to uncover insights from the data.

## Shiny App

# Section 2: Predictive Modeling
This section will involve building and evaluating classification models.
## Classification
# Section 3: Clustering
This section will involve performing clustering analysis to discover patterns in the data.

# Conclusion
Summarize the key findings and insights from your analysis.
